---
title: "4_land_use_changes — PLUM Alexander (id99, HILDA+ entities, p05 GeoTIFFs)"
output: html_notebook
---

```{r setup, message=FALSE}
library(terra)
library(tidyterra)
library(tidyverse)
library(ggplot2)
library(stringr)
library(sf)
library(rnaturalearth)

if (requireNamespace("rstudioapi", quietly = TRUE) &&
    rstudioapi::isAvailable()) {
  setwd(dirname(rstudioapi::getSourceEditorContext()$path))
}
```

# 0. Configuration

Unlike the model-agnostic notebook, PLUM Alexander (id99) is read
directly from the aggregated 0.5° GeoTIFFs produced by the coarsening
notebook (`...PLUM-Alexander__coarsen-to-0.5deg...Rmd`), not from an
EBVcube NetCDF. This is because the native cube is binary (0/1) at
0.01° and the coarsened proportional-cover TIFFs are the correct input
for the cSAR pipeline.

The entity → target class mapping mirrors the YAML contract used by
`4_land_use_changes_agnostic_3.Rmd` for the other models.

```{r config}
model_label <- "PLUM_Alexander"

# Root of the aggregated 0.5° GeoTIFFs
tif_root <- "../preprocessed/PLUM_alexander/0.5deg"

scenarios <- c("NfN", "NfNl", "NaC", "NfS", "BAU")
years     <- c(2020, 2030, 2050, 2070)
n_entities <- 7

# Entity names in the exact order stored in the cube / output by the
# coarsening notebook (entity_01 ... entity_07)
entity_names <- c(
  "Urban",
  "Cropland",
  "Pasture",
  "Forest managed",
  "Forest unmanaged",
  "Grass/shrubland",
  "Other land"
)

# Safe names for list keys (spaces and slashes → underscore)
safe_entity_names <- gsub("[[:space:]/]+", "_", entity_names)

# Entity → target class mapping (matches YAML logic for other models)
# NA = excluded from habitat-area denominator
category_map <- c(
  Urban              = "Urban",
  Cropland           = "Cropland",
  Pasture            = "Pasture",
  Forest_managed     = "Secondary_vegetation",
  Forest_unmanaged   = "Primary_vegetation",
  Grass_shrubland    = "Primary_vegetation",
  Other_land         = "Primary_vegetation"
)

# No NA-mapped entities for PLUM Alexander — all 7 contribute
category_map_active <- category_map[!is.na(category_map)]
new_categories      <- unique(category_map_active)

cat("Model:         ", model_label, "\n")
cat("Scenarios:     ", paste(scenarios, collapse = ", "), "\n")
cat("Years:         ", paste(years, collapse = ", "), "\n")
cat("Target classes:", paste(new_categories, collapse = ", "), "\n")
```

# 1. Paths and spatial template

The 0.5° global template is derived from the first available GeoTIFF
so that `cellSize()` uses the correct extent and resolution — identical
to the approach used in the agnostic notebook via `cube_props@spatial`.

```{r paths}
stopifnot(dir.exists(tif_root))

# IPBES regions shapefile — same discovery logic as agnostic notebook
ipbes_shp <- list.files("../Data_geo", pattern = "\\.shp$",
                         full.names = TRUE, ignore.case = TRUE,
                         recursive  = TRUE)
ipbesred <- vect(ipbes_shp[1])

# Spatial template from the first available p05 TIFF
template_tif <- file.path(tif_root, scenarios[1],
                           years[1], "entity_01_p05.tif")
stopifnot(file.exists(template_tif))
r_template <- rast(template_tif)

# Pixel-area raster (latitude-varying), built once — matches agnostic notebook
pixel_area_km2 <- cellSize(r_template, unit = "km")

cat("Spatial template:\n")
print(r_template)
```

# 2. Helper: read all entities for one scenario and two timesteps

Mirrors the `ebv_read()` loop in section 4.1 of the agnostic notebook,
but reads from p05 GeoTIFFs instead of an EBVcube NetCDF. Each entity
is read once and stored in `lu_frac` — reused for both the
percentage-point and km² calculations below.

```{r read-helper}
read_lu_frac <- function(scenario, t0, t1) {
  lu_frac <- vector("list", n_entities)
  names(lu_frac) <- safe_entity_names

  for (e in seq_len(n_entities)) {
    tif_t0 <- file.path(tif_root, scenario, t0,
                        sprintf("entity_%02d_p05.tif", e))
    tif_t1 <- file.path(tif_root, scenario, t1,
                        sprintf("entity_%02d_p05.tif", e))

    stopifnot(file.exists(tif_t0), file.exists(tif_t1))

    r <- c(rast(tif_t0), rast(tif_t1))
    names(r) <- paste0(safe_entity_names[e], "_", c(t0, t1))

    lu_frac[[safe_entity_names[e]]] <- r
    cat(sprintf("[%d/%d] %s\n", e, n_entities, entity_names[e]))
  }

  lu_frac
}
```

# 3. Select scenario and timesteps

Change `scenario_choice`, `t0_label`, and `t1_label` to switch
scenarios or time windows — mirrors the `scenario_choice` / `t0_label`
/ `t1_label` pattern in the agnostic notebook.

```{r scenario-select}
scenario_choice <- "NfS"   # one of: NfN, NfNl, NaC, NfS, BAU
t0_label        <- "2020"
t1_label        <- "2050"

stopifnot(scenario_choice %in% scenarios)
stopifnot(t0_label %in% as.character(years))
stopifnot(t1_label %in% as.character(years))

cat("Processing scenario:", scenario_choice, "\n")
cat("Baseline (t0):      ", t0_label, "\n")
cat("Target  (t1):       ", t1_label, "\n")
```

# 4. Changes in area

## 4.1 Read fractions and compute percentage-point change per decade

```{r fractions}
lu_frac <- read_lu_frac(scenario_choice, t0_label, t1_label)

cat("Done.", model_label, "-", n_entities, "entities x 2 timesteps.\n")

# Recategorise: sum fractions across entities sharing a target class
lu_frac_recat <- lapply(new_categories, function(new_cat) {
  old_entities  <- names(category_map_active)[category_map_active == new_cat]
  summed        <- Reduce(`+`, lu_frac[old_entities])
  names(summed) <- paste0(new_cat, "_", c(t0_label, t1_label))
  summed
}) |> setNames(new_categories)

cat("Recategorised into", length(new_categories), "classes:\n")
print(new_categories)

# Percentage-point change per decade
n_decades <- (as.numeric(t1_label) - as.numeric(t0_label)) / 10

lu_pct_change_per_decade <- lapply(names(lu_frac_recat), function(lu) {
  r  <- lu_frac_recat[[lu]]
  y0 <- r[[paste0(lu, "_", t0_label)]]
  y1 <- r[[paste0(lu, "_", t1_label)]]
  change_per_decade <- (y1 - y0) / n_decades
  names(change_per_decade) <-
    paste0(lu, "_pctchange_per_decade_", t0_label, "_", t1_label)
  change_per_decade
}) |> setNames(names(lu_frac_recat))

cat("Percentage-point change per decade computed for", model_label, ":\n")
print(names(lu_pct_change_per_decade))

plot(lu_pct_change_per_decade$Primary_vegetation,
     main = paste(model_label, "- Primary_vegetation pct change/decade"))
```

```{r pct-change-stack}
pct_change_stack <- rast(lu_pct_change_per_decade)
plot(pct_change_stack, main = paste(model_label, names(pct_change_stack)))
```

## 4.2 Changes in km²

Fractions are converted to km² using the latitude-varying pixel-area
raster built in section 1 — no second read of the TIFFs, identical to
the agnostic notebook.

```{r km2}
lu_area_km2 <- lapply(lu_frac, function(r) r * pixel_area_km2)
names(lu_area_km2) <- names(lu_frac)

cat("Converted to km² for", model_label, "-", length(lu_area_km2), "entities.\n")

# Recategorise km²
lu_area_km2_recat <- lapply(new_categories, function(new_cat) {
  old_entities  <- names(category_map_active)[category_map_active == new_cat]
  summed        <- Reduce(`+`, lu_area_km2[old_entities])
  names(summed) <- paste0(new_cat, "_", c(t0_label, t1_label))
  summed
}) |> setNames(new_categories)

cat("Reclassified into", length(new_categories), "categories for",
    model_label, ":\n")
print(new_categories)
```

```{r km2-plot}
lu_area_stack <- rast(lu_area_km2_recat)
plot(lu_area_stack, main = paste(model_label, names(lu_area_stack)))

# Primary_vegetation: t0, t1, and difference
pv_t0   <- lu_area_km2_recat$Primary_vegetation[[paste0("Primary_vegetation_", t0_label)]]
pv_t1   <- lu_area_km2_recat$Primary_vegetation[[paste0("Primary_vegetation_", t1_label)]]
pv_diff <- pv_t1 - pv_t0
names(pv_diff) <- paste0("Primary_vegetation_diff_", t0_label, "_", t1_label)

pv_compare <- c(pv_t0, pv_t1, pv_diff)
plot(pv_compare, main = paste(model_label, "-", names(pv_compare)))
```

# 5. Plots

## 5.1 Map: percentage-point change per decade (continuous and binned)

```{r maps}
coastline <- ne_coastline(scale = "medium", returnclass = "sf")
coastline <- st_make_valid(coastline)

pv_pct_change <- lu_pct_change_per_decade$Primary_vegetation
pv_df <- as.data.frame(pv_pct_change, xy = TRUE, na.rm = TRUE)
names(pv_df)[3] <- "pct_change"

# Alternative A: continuous scale
ggplot(pv_df) +
  geom_tile(aes(x = x, y = y, fill = pct_change)) +
  geom_sf(data = coastline, color = "grey30", linewidth = 0.1, inherit.aes = FALSE) +
  scale_fill_distiller(
    palette   = "Reds",
    direction = -1,
    limits    = c(min(pv_df$pct_change, na.rm = TRUE), 0),
    oob       = scales::squish,
    na.value  = NA,
    name      = "pp / decade"
  ) +
  coord_sf(expand = FALSE, xlim = c(-180, 180), ylim = c(-90, 90)) +
  labs(title = paste0(model_label, " — Δ Primary_vegetation ", t0_label, "–", t1_label)) +
  theme_void() +
  theme(
    legend.position   = "bottom",
    legend.key.width  = unit(2, "cm"),
    legend.key.height = unit(0.3, "cm")
  )

# Alternative B: binned scale
breaks_b   <- c(min(pv_df$pct_change, na.rm = TRUE),
                -3.08, -1.16, -0.32, -0.30, -0.2, -0.1, -0.05, -0.01, 0)
reds_colors <- rev(RColorBrewer::brewer.pal(9, "Reds"))
reds_colors[length(reds_colors)] <- "#FFF7BC"

ggplot(pv_df) +
  geom_tile(aes(x = x, y = y, fill = pct_change)) +
  geom_sf(data = coastline, color = "grey30", linewidth = 0.1, inherit.aes = FALSE) +
  scale_fill_stepsn(
    colors   = reds_colors,
    breaks   = breaks_b,
    labels   = round(sort(breaks_b), 2),
    na.value = NA,
    name     = "percentage points per decade"
  ) +
  coord_sf(expand = FALSE, xlim = c(-180, 180), ylim = c(-90, 90)) +
  labs(title = paste0(model_label, " — Δ ", t0_label, "–", t1_label)) +
  theme_void() +
  theme(
    legend.position   = "bottom",
    legend.key.width  = unit(1.4, "cm"),
    legend.key.height = unit(0.3, "cm")
  )
```

```{r binned-diagnostic}
breaks_b <- c(min(pv_df$pct_change, na.rm = TRUE),
              -3.08, -1.16, -0.32, -0.30, -0.2, -0.1, -0.05, -0.01, 0)
table(cut(pv_df$pct_change, breaks = sort(unique(breaks_b))))
```

## 5.2 Fraction of natural land in baseline year (t0)

```{r natural-land-distribution}
pv_pct_2020 <- lu_frac_recat$Primary_vegetation[[paste0("Primary_vegetation_", t0_label)]] * 100
names(pv_pct_2020) <- paste0("Primary_vegetation_pct_", t0_label)

pv_pct_df <- as.data.frame(pv_pct_2020, xy = TRUE, na.rm = TRUE)
names(pv_pct_df)[3] <- "pct"

probs_dist  <- c(0, 0.05, 0.15, 0.3, 0.45, 0.55, 0.7, 0.85, 0.95, 1)
breaks_dist <- unname(quantile(pv_pct_df$pct, probs = probs_dist, na.rm = TRUE))

print(table(cut(pv_pct_df$pct, breaks = sort(unique(breaks_dist)))))

brbg_colors <- RColorBrewer::brewer.pal(length(unique(breaks_dist)) - 1, "BrBG")

ggplot(pv_pct_df) +
  geom_tile(aes(x = x, y = y, fill = pct)) +
  geom_sf(data = coastline, color = "grey30", linewidth = 0.1, inherit.aes = FALSE) +
  scale_fill_stepsn(
    colors   = brbg_colors,
    breaks   = breaks_dist,
    labels   = round(breaks_dist, 1),
    limits   = c(0, 100),
    na.value = NA,
    name     = paste0("distribution of land in ", t0_label, " (%)")
  ) +
  coord_sf(expand = FALSE, xlim = c(-180, 180), ylim = c(-90, 90)) +
  labs(title = paste0(model_label, " — ", t0_label)) +
  theme_void() +
  theme(
    legend.position    = "bottom",
    legend.key.width   = unit(1.4, "cm"),
    legend.key.height  = unit(0.3, "cm"),
    plot.title         = element_text(size = 14, face = "bold", hjust = 0)
  )
```

```{r quantile-diagnostic}
breaks_dist <- unname(quantile(pv_pct_df$pct, probs = probs_dist, na.rm = TRUE))
breaks_dist
table(cut(pv_pct_df$pct, breaks = sort(unique(breaks_dist))))
```
