---
title: "PLUM (Alexander/LandSyMM) Land Cover - Coarsen to 0.5deg and Build EBV Cube"
author: "Andres Marmol Guijarro"
date: "2026-06-21"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

# 1. Introduction

This workflow takes the native-resolution (0.01°, binary 0/1) PLUM
land-cover GeoTIFFs produced by the extraction notebook
(`...PLUM-Alexander__extract-tiffs-from-NetCDF...Rmd`) and aggregates
them to a coarser **0.5° resolution**, then writes the result into a
new EBV Cube NetCDF — following the same structure used in the GLOBIO
downscaling/aggregation workflow.

The aggregation method matches what was used for GLOBIO: for each
0.5° grid cell, the workflow **counts the number of native 0.01°
pixels** that fall within it and **counts how many of those pixels
are land-cover = 1**, then derives the *proportion of pixels with
that land cover* as `n_pixels_landcover / n_pixels_total`. This is
computed efficiently via `terra::aggregate(fun = mean)` on the binary
layer, since the mean of a 0/1 layer is mathematically identical to
that pixel proportion — but expressed here as an explicit pixel-count
ratio for clarity.

Two core steps:

1.  *Aggregate the native 0.01° rasters to 0.5°*\
    The original PLUM/HILDA+ rasters have a spatial resolution of
    0.01° (~1 km at the equator). Each 0.5° output cell is computed as
    the proportion of native pixels classified as that land-cover
    entity within the coarse cell footprint.

2.  *Build and populate a new EBV Cube NetCDF at 0.5°*\
    The aggregated rasters are inserted into a new EBV-compliant
    NetCDF, structured by scenario, metric, time step, and entity —
    matching the structure of the source cube
    (`alexander_ecostr_id99_20260217.nc`).

The 0.5° raster template explicitly spans the **full global extent**
(-180/180 lon, -90/90 lat), matching the GLOBIO 0.5° template and
ensuring Antarctica is retained even if some native input tiles are
sparse or all-NA at high southern latitudes.

------------------------------------------------------------------------

## 1.1 PLUM scenarios and time steps

| **Scenario** | **Abbreviation** | **Available Years** |
|---|---|---|
| Business as Usual SSP2 RCP4.5 | BAU | 2020, 2030, 2050, 2070 |
| Nature as Culture RCP4.5 | NaC | 2020, 2030, 2050, 2070 |
| Nature for Society RCP4.5 | NfS | 2020, 2030, 2050, 2070 |
| Nature for Nature RCP4.5 | NfN | 2020, 2030, 2050, 2070 |
| Nature for Nature (local) RCP4.5 | NfNl | 2020, 2030, 2050, 2070 |

## 1.2 Entities (HILDA+ harmonised classes)

7 of the 10 original HILDA+ classes (oceans, water, and no-data
excluded):

1.  urban
2.  cropland
3.  pasture
4.  forest managed
5.  forest unmanaged
6.  grass/shrubland
7.  other land

## 1.3 Before you start

-   The extraction notebook has already been run, and native-resolution
    `entity_XX_native.tif` files exist under `0.01deg_native/<scenario>/<year>/`.
-   The `metadata.json` file for this EBV dataset exists.
-   Sufficient temporary disk space is available for raster aggregation.
-   The `ebvcube` package is installed and working correctly.

------------------------------------------------------------------------

## 2. Load libraries

```{r libraries}
rm(list = ls())
gc()

library(terra)
library(ebvcube)
library(here)
library(stringr)
library(dplyr)
```

------------------------------------------------------------------------

## 3. Configure temporary raster storage

```{r temp-storage}
local_temp <- "D:/temp"
dir.create(local_temp, showWarnings = FALSE, recursive = TRUE)

terraOptions(
  todisk  = TRUE,
  memfrac = 0.3,
  tempdir = local_temp
)

cat("Local temporary folder:", local_temp, "\n")
```

------------------------------------------------------------------------

## 4. Define input and output paths

```{r paths}
in_root  <- here("preprocessed", "PLUM_alexander", "0.01deg_native")
out_root <- here("preprocessed", "PLUM_alexander", "0.5deg")
dir.create(out_root, showWarnings = FALSE, recursive = TRUE)

cat("Input root: ", in_root, "\n")
cat("Output root:", out_root, "\n")
```

------------------------------------------------------------------------

## 5. Define workflow settings

```{r settings}
id        <- 99
years     <- c(2020, 2030, 2050, 2070)
scenarios <- c("NfN", "NfNl", "NaC", "NfS", "BAU")

entities <- c(
  "urban",
  "cropland",
  "pasture",
  "forest managed",
  "forest unmanaged",
  "grass/shrubland",
  "other land"
)

n_entities <- length(entities)

cat("ID:", id, "\n")
cat("Years:", paste(years, collapse = ", "), "\n")
cat("Scenarios:", paste(scenarios, collapse = ", "), "\n")
cat("Entities:", paste(entities, collapse = ", "), "\n")
```

------------------------------------------------------------------------

## 6. Build the 0.5° global raster template

This template defines the target grid for aggregation. It is forced to
the **full global extent** (-180/180, -90/90) regardless of the input
data's actual extent, so Antarctica and any sparse/empty high-latitude
tiles are preserved as NA rather than silently dropped.

```{r template}
target_deg <- 0.5

r_template <- rast(
  xmin = -180, xmax = 180,
  ymin = -90,  ymax = 90,
  resolution = target_deg,
  crs = "EPSG:4326"
)

r_template

cube_extent <- as.vector(ext(r_template))
cube_res    <- res(r_template)
epsg        <- 4326

cat("Target resolution:", target_deg, "degrees\n")
cat("Target extent (xmin, xmax, ymin, ymax):", cube_extent, "\n")
```

------------------------------------------------------------------------

# 7. Aggregation to 0.5° resolution

For each scenario, year, and entity, this section:

1.  reads the native 0.01° binary GeoTIFF,
2.  aggregates it to 0.5° using the **mean** of the binary 0/1 values
    within each coarse cell — i.e. `n_pixels_landcover / n_pixels_total`,
3.  resamples/matches the result onto the fixed global 0.5° template so
    every output raster shares an identical extent (including
    Antarctica), and
4.  writes the result as a proportional-cover GeoTIFF.

```{r aggregation-loop}
total_iter   <- length(scenarios) * length(years) * n_entities
current_iter <- 0
pb <- txtProgressBar(min = 0, max = total_iter, style = 3)
t_start <- proc.time()

created <- character()
errors  <- character()

for (sc in scenarios) {

  for (yr in years) {

    in_dir  <- file.path(in_root, sc, as.character(yr))
    out_dir <- file.path(out_root, sc, as.character(yr))
    dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

    cat("\n==========", sc, yr, "==========\n")

    for (e_no in seq_len(n_entities)) {

      tif_in <- file.path(in_dir, sprintf("entity_%02d_native.tif", e_no))

      if (!file.exists(tif_in)) {
        msg <- paste("Missing native file:", tif_in)
        errors <- c(errors, msg)
        warning(msg)
        current_iter <- current_iter + 1
        setTxtProgressBar(pb, current_iter)
        next
      }

      res_agg <- tryCatch({

        r <- rast(tif_in)

        # fact = how many native pixels fall along each axis of one
        # 0.5 deg coarse cell (native res is ~0.01 deg)
        fact <- target_deg / res(r)[1]

        # Aggregate - count of land-cover pixels / total pixels per
        # coarse cell, via mean() of the binary (0/1) layer
        r_d <- aggregate(
          r,
          fact   = fact,
          fun    = mean,
          na.rm  = TRUE
        )

        # Snap/extend onto the fixed global template so every output
        # raster shares an identical grid, including Antarctica
        r_d <- extend(r_d, r_template)
        r_d <- resample(r_d, r_template, method = "near")

        names(r_d) <- entities[e_no]

        out_file <- file.path(out_dir, sprintf("entity_%02d_p05.tif", e_no))

        writeRaster(
          r_d,
          out_file,
          datatype  = "FLT4S",
          overwrite = TRUE,
          wopt      = list(gdal = c("COMPRESS=DEFLATE"))
        )

        rm(r, r_d)
        gc()
        out_file

      }, error = function(e) e)

      if (inherits(res_agg, "error")) {
        msg <- paste("FAILED | scenario:", sc, "year:", yr,
                     "entity:", entities[e_no], "|", conditionMessage(res_agg))
        errors <- c(errors, msg)
        warning(msg)
      } else {
        created <- c(created, res_agg)
      }

      current_iter <- current_iter + 1
      setTxtProgressBar(pb, current_iter)
    }
  }
}

close(pb)
t_total <- (proc.time() - t_start)["elapsed"]

cat("\nAggregation completed in", round(t_total / 60, 2), "minutes.\n")
cat("Created:", length(created), "\n")
cat("Failed: ", length(errors),  "\n")
if (length(errors) > 0) print(head(errors, 10))
```

------------------------------------------------------------------------

## 7.1 Diagnostic: confirm aggregation matches expectation

```{r diagnostic-agg, eval=FALSE}
check_file <- file.path(out_root, "BAU", "2020", "entity_01_p05.tif")

r_check <- rast(check_file)
r_check
plot(r_check, main = "BAU 2020 - urban proportional cover (0.5 deg)")

cat("Value range:", paste(round(range(values(r_check, mat = FALSE), na.rm = TRUE), 4),
                          collapse = " to "), "\n")
cat("Extent matches template:", identical(ext(r_check), ext(r_template)), "\n")
```

------------------------------------------------------------------------

# 8. Create the EBV Cube NetCDF file

## 8.1 Define paths

```{r nc-paths}
creation_date <- format(Sys.Date(), "%Y%m%d")
cat("Created date:", creation_date, "\n")
cat("ID:", id, "\n")

json_root <- here("json")
js <- file.path(json_root, paste0("metadata_", id, ".json"))

# Naming convention: "<first_author>_<ebv_class>_id<ID>_<YYYYMMDD>.nc"
nc_out <- here("Data_geo", "land_use", "PLUM", "cubes",
               paste0("alexander_ecostr_id", id, "_", creation_date, "_p05.nc"))

cat("EBV Cube NetCDF file name:", basename(nc_out), "\n")
cat("Aggregated raster root:   ", out_root, "\n")

stopifnot(file.exists(js))

tif_files <- list.files(out_root, pattern = "\\.tif$", recursive = TRUE, full.names = TRUE)
cat("Total aggregated GeoTIFF files found:", length(tif_files), "\n")
```

## 8.2 Define EBV Cube structure

```{r nc-structure}
cat("Entity variable\n")
cat("=======================\n")
cat("Entity names:", paste(entities, collapse = ", "), "\n")
cat("Entities number:", n_entities, "\n")

# Spatial settings come from the fixed global template (Section 6),
# not from any single input raster, so the cube always covers the
# full globe including Antarctica
list(
  timesteps = years,
  extent    = cube_extent,
  res       = cube_res,
  epsg      = epsg
)

# Fill value and datatype (proportional cover, not binary)
fillvalue <- -3.4e+38
prec      <- "float"
```

## 8.3 Create empty NetCDF file

```{r create-cube}
timesteps <- as.character(years)

ebv_create(
  jsonpath   = js,
  outputpath = nc_out,
  entities   = entities,
  epsg       = epsg,
  extent     = cube_extent,
  resolution = cube_res,
  timesteps  = timesteps,
  fillvalue  = fillvalue,
  prec       = prec,
  force_4D   = TRUE,
  overwrite  = TRUE,
  verbose    = TRUE
)

# QC: inspect cube structure - should show 5 scenarios x 1 metric x 7 entities
dc_new <- ebv_datacubepaths(nc_out, verbose = FALSE)
print(dc_new)

# QC: spatial properties
prop <- ebv_properties(nc_out, scenario = 1, metric = 1, verbose = FALSE)
print(prop@spatial)
```

> The JSON file must define the entities and metric exactly as listed
> above. Verify this before running `ebv_create()`.

------------------------------------------------------------------------

# 9. Populate the EBV Cube

```{r insert-data}
metric_no <- 1

start_time <- Sys.time()

for (s_no in seq_along(scenarios)) {

  sc <- scenarios[s_no]

  for (t_no in seq_along(years)) {

    yr <- years[t_no]
    in_dir <- file.path(out_root, sc, as.character(yr))

    cat("\n=== INSERT | scenario:", sc, "| year:", yr, "===\n")

    if (!dir.exists(in_dir)) {
      warning("Input directory does not exist: ", in_dir)
      next
    }

    for (e_no in seq_len(n_entities)) {

      tif_in <- file.path(in_dir, sprintf("entity_%02d_p05.tif", e_no))

      if (!file.exists(tif_in)) {
        warning("Missing aggregated file: ", tif_in)
        next
      }

      ebv_add_data(
        filepath_nc = nc_out,
        data        = tif_in,
        scenario    = s_no,
        metric      = metric_no,
        entity      = e_no,
        timestep    = t_no,
        ignore_RAM  = FALSE,
        raw         = FALSE,
        verbose     = FALSE
      )

      gc(verbose = FALSE)
      cat("  inserted entity", e_no, "(", entities[e_no], ")\n")
    }
  }
}

end_time   <- Sys.time()
time_taken <- end_time - start_time
print(time_taken)

cat("\nDONE. All aggregated entities inserted into:\n", nc_out, "\n")
```

------------------------------------------------------------------------

## 10. Final check

```{r final-check, eval=FALSE}
mycube <- ebv_read(
  filepath     = nc_out,
  datacubepath = NULL,
  entity       = "urban",
  timestep     = 1,
  type         = "r",
  scenario     = "BAU",
  metric       = 1
)

terra::plot(mycube, main = "BAU 2020 - urban proportional cover (0.5 deg)")

ebv_properties(nc_out)
ebv_entities(nc_out)
ebv_timesteps(nc_out)
```

------------------------------------------------------------------------

## Common errors

| Error | Likely cause | Fix |
|---|---|---|
| Missing native file warnings | Extraction notebook not run, or `id`/path mismatch | Run extraction notebook first; verify `in_root` |
| Extent mismatch after aggregation | Native tile doesn't span the full globe | `extend()` + `resample()` step in Section 7 handles this onto the fixed template |
| `ebv_create()` fails on entities | JSON metadata entity list doesn't match `entities` vector exactly | Check `metadata_<id>.json` entity names/order |
| Aggregated values outside [0,1] | Native raster isn't strictly binary 0/1 | Check `entity_XX_native.tif` values before aggregating |
