---
title: "PLUM (Alexander/LandSyMM) Land Cover - Extract GeoTIFFs from EBV NetCDF"
author: "Andres Marmol Guijarro"
date: "2026-06-21"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

# Overview

This notebook extracts all entity / scenario / timestep slices from the
PLUM (LandSyMM-PLUM, Peter Alexander) EBV Cube NetCDF
(`alexander_ecostr_id99_20260217.nc`) and writes each one out as a
single-band GeoTIFF at native resolution (0.01°, ~1 km).

The source cube classifies land cover according to the **HILDA+**
classification (harmonised to 7 of the 10 original HILDA+ classes):
urban, cropland, pasture, forest managed, forest unmanaged,
grass/shrubland, and other land. Values are binary (0/1) per pixel —
each entity layer records whether that land-cover type occurs in a
given cell, for a given scenario and timestep.

Output naming follows the same `entity_XX` convention used in the
GLOBIO downscaling notebook, so the next notebook
(`...coarsen-to-0.5deg...Rmd`) can reuse the same aggregation logic
with minimal changes:

```
<out_root>/<scenario>/<year>/entity_<NN>_native.tif
```

------------------------------------------------------------------------

## 1. Load libraries

```{r libraries}
rm(list = ls())
gc()

library(terra)
library(here)
library(stringr)
library(ebvcube)

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

------------------------------------------------------------------------

## 2. Configure temporary raster storage

```{r temp-storage}
local_temp <- "../temp"
dir.create(local_temp, showWarnings = FALSE, recursive = TRUE)

terraOptions(
  todisk  = TRUE,
  memfrac = 0.3,
  tempdir = local_temp
)

cat("Local temporary folder:", local_temp, "\n")
```

------------------------------------------------------------------------

## 3. Define paths

```{r paths}
# Path to the source EBV Cube NetCDF
nc_in <- here("Data_geo", "land_use", "PLUM", "cubes",
              "alexander_ecostr_id99_20260217.nc")

stopifnot(file.exists(nc_in))

# Root folder where extracted native-resolution GeoTIFFs will be written
out_root <- here("preprocessed", "PLUM_alexander", "0.01deg_native")
dir.create(out_root, showWarnings = FALSE, recursive = TRUE)

cat("Input NetCDF:", nc_in, "\n")
cat("Output root: ", out_root, "\n")
```

------------------------------------------------------------------------

## 4. Inspect cube structure

Before extracting, confirm the scenario, entity, and time structure of
the cube. This avoids hard-coding values that may not match the file.

```{r inspect}
library(ebvcube)

cubepaths <- ebv_datacubepaths(nc_in, verbose = FALSE)
print(cubepaths)

props <- ebv_properties(nc_in, scenario = 1, metric = 1, verbose = FALSE)
print(props@spatial)

entities  <- props@general$entity_names
timesteps <- props@temporal$timesteps

cat("Entities: ",  paste(entities, collapse = ", "), "\n")
cat("Timesteps:",  paste(timesteps, collapse = ", "), "\n")
cat("Scenarios:",  paste(unique(cubepaths$scenario_names), collapse = ", "), "\n")
```

------------------------------------------------------------------------

## 5. Define scenarios, entities, and years to extract

These are read from the cube directly in the previous step, but pinned
here explicitly so the extraction loop is transparent and reproducible
even if the cube changes later.

```{r settings}
scenarios <- c("NfN", "NfNl", "NaC", "NfS", "BAU")

# Entity order as it appears in the NetCDF (HILDA+ harmonised classes)
entities <- c(
  "Urban",
  "Cropland",
  "Pasture",
  "Forest managed",
  "Forest unmanaged",
  "Grass/shrubland",
  "Other land"
)

# Years available in the cube (decadal timesteps, 2020-2070)
years <- c("2020", "2030", "2050", "2070")

cat("Scenarios:", paste(scenarios, collapse = ", "), "\n")
cat("Entities: ", paste(entities, collapse = ", "), "\n")
cat("Years:    ", paste(years, collapse = ", "), "\n")
```

------------------------------------------------------------------------

# 6. Extraction loop

For each scenario, timestep, and entity, this reads the corresponding
2D slice from the NetCDF via `ebv_read()` and writes it out as a
single-band native-resolution GeoTIFF.

```{r extraction-loop}
total_iter   <- length(scenarios) * length(years) * length(entities)
current_iter <- 0
pb <- txtProgressBar(min = 0, max = total_iter, style = 3)
t_start <- proc.time()

created <- character()
errors  <- character()

for (sc in scenarios) {

  for (yr in years) {

    out_dir <- file.path(out_root, sc, yr)
    dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

    for (e_no in seq_along(entities)) {

      ent <- entities[e_no]

      out_file <- file.path(
        out_dir,
        sprintf("entity_%02d_native.tif", e_no)
      )

      res_extract <- tryCatch({

        r <- ebv_read(
          filepath     = nc_in,
          datacubepath = NULL,
          entity       = ent,
          timestep     = which(years == yr),
          type         = "r",
          scenario     = sc,
          metric       = 1
        )

        names(r) <- ent

        writeRaster(
          r,
          out_file,
          datatype  = "INT1U",
          overwrite = TRUE,
          wopt      = list(gdal = c("COMPRESS=DEFLATE"))
        )

        rm(r)
        gc()
        out_file

      }, error = function(e) e)

      if (inherits(res_extract, "error")) {
        msg <- paste("FAILED | scenario:", sc, "year:", yr,
                     "entity:", ent, "|", conditionMessage(res_extract))
        errors <- c(errors, msg)
        warning(msg)
      } else {
        created <- c(created, res_extract)
      }

      current_iter <- current_iter + 1
      setTxtProgressBar(pb, current_iter)
    }
  }
}

close(pb)
t_total <- (proc.time() - t_start)["elapsed"]

cat("\nExtraction completed in", round(t_total / 60, 2), "minutes.\n")
cat("Created:", length(created), "\n")
cat("Failed: ", length(errors),  "\n")
if (length(errors) > 0) print(head(errors, 10))
```

------------------------------------------------------------------------

## 7. Expected output structure

```
0.01deg_native/
├── NfN/
│   ├── 2020/
│   │   ├── entity_01_native.tif   # urban
│   │   ├── entity_02_native.tif   # cropland
│   │   ├── ...
│   │   └── entity_07_native.tif   # other land
│   ├── 2030/
│   └── ...
├── NfNl/
├── NaC/
├── NfS/
└── BAU/
```

Each `.tif` is single-band, binary (0/1), at the native resolution of
the source cube (0.01°, EPSG:4326).

------------------------------------------------------------------------

## 8. Diagnostic: spot-check one extracted raster

Run this independently to confirm an extracted raster matches the
source cube.

```{r diagnostic, eval=FALSE}
check_file <- file.path(out_root, "BAU", "2020", "entity_01_native.tif")

r_check <- rast(check_file)
r_check
plot(r_check, main = "BAU 2020 - urban (entity_01)")

cat("Unique values:", paste(unique(values(r_check, mat = FALSE)), collapse = ", "), "\n")
```

------------------------------------------------------------------------

## Common errors

| Error | Likely cause | Fix |
|---|---|---|
| `file.exists(nc_in)` fails | SSD/network drive not mounted, or path changed | Verify `/Volumes/AM_SSD/...` is mounted; adjust `here()` root |
| `ebv_read()` entity not found | Entity name string doesn't match cube exactly | Check `ebv_entities(nc_in)` for exact spelling/case |
| Missing `rhdf5` package | Bioconductor dependency not installed | `BiocManager::install("rhdf5")` |
| Slow extraction | Reading from spinning/network disk | Copy `.nc` to local SSD before running |
