---
title: "4_land_use_changes (agnostic, config-driven)"
output: html_notebook
---

```{r setup, message=FALSE}
library(ebvcube)
library(tidyterra)
library(terra)
library(tidyverse)
library(graticule)
library(sf)
library(rworldmap)
library(ggplot2)
library(stringr)
library(yaml)

# Anchor wd to the notebook location (interactive use only;
# knitr sets root.dir to the document directory automatically)
if (requireNamespace("rstudioapi", quietly = TRUE) &&
    rstudioapi::isAvailable()) {
  setwd(dirname(rstudioapi::getSourceEditorContext()$path))
}
```

# 0. Configuration

The per-model contract — cube/file location, entity → target class mapping,
and which EBVcube metric to consume — lives in
`../config/category_map.yaml`. Switching models is a single line: set
`model_key` below to any key declared under `models:` in the YAML. The
mapping and metric filter follow automatically.

Two input backends are supported, declared per-model via `input_type` in
the YAML:

- `"ebvcube"` (default if unset) — reads an EBVcube NetCDF via
  `ebvcube::ebv_read()`. Used by `GLOBIO`, `PLUM`, `ScenMIP`.
- `"geotiff_dir"` — reads a folder of pre-processed, per-scenario/per-year/
  per-entity GeoTIFFs via `terra::rast()`. Used by `PLUM_harmonised`
  (id99, HILDA+-harmonised, coarsened to 0.5°) — this was previously
  handled by its own notebook, `5_land_use_changes_PLUM_Alexander.Rmd`
  (now archived), folded in here 2026-09-03.

**TODO / TEMPORARY:** `geotiff_dir` is a stopgap, not the target design. The intent is for every model to ultimately be read from a proper EBV data cube — `PLUM_harmonised` uses `geotiff_dir` only because to accelerate reading processes to provide preliminary outputs for BES-SIM 2 in person meeting 2026. It remains to switch this model to `input_type: ebvcube` in the YAML and the `geotiff_dir` branches below (sections 1, 2, 4.1) can be removed.

Everything downstream of section 4.1 (recategorization, percentage-point
change, km² conversion, plotting) is identical regardless of backend — only
the "read fractions for this model" step differs.

```{r config}
cfg <- yaml::read_yaml("../config/category_map.yaml")

# Single switch: pick the model whose cube to process
model_key   <- "GLOBIO"               # one of names(cfg$models)
model_label <- model_key

stopifnot(model_key %in% names(cfg$models))
mcfg <- cfg$models[[model_key]]

# Input backend: "ebvcube" (default) or "geotiff_dir" — see note above
input_type <- if (!is.null(mcfg$input_type)) mcfg$input_type else "ebvcube"

cat("Active model:   ", model_label, "\n")
cat("Input type:     ", input_type, "\n")
cat("Config version: ", cfg$version, "\n")
```

# 1. Paths

For `ebvcube` models, the cube file path resolves as
`paths$data_root + models[[model]]$file` from the YAML. For `geotiff_dir`
models, `models[[model]]$tif_root` points directly at the folder of
coarsened GeoTIFFs. The IPBES regions shapefile is discovered by recursive
search under `data_root` either way (using the pre-simplified version
produced previously with `simplifyGeom(tolerance = 0.05)`).

```{r paths}
root_LU_cubes <- cfg$paths$data_root

if (input_type == "ebvcube") {

  cube_path <- file.path(root_LU_cubes, mcfg$file)
  stopifnot(file.exists(cube_path))

} else if (input_type == "geotiff_dir") {

  # TEMPORARY (see TODO in section 0): GeoTIFF path, not an EBVcube read.
  tif_root <- mcfg$tif_root
  stopifnot(dir.exists(tif_root))

} else {
  stop("Unknown input_type '", input_type, "' for model '", model_key, "'")
}

ipbes_shp <- list.files(root_LU_cubes, pattern = "\\.shp$",
                        full.names = TRUE, ignore.case = TRUE,
                        recursive  = TRUE)

# Reduced (pre-simplified) shape — produced once with:
# ipbesred <- simplifyGeom(vect(ipbes_full), tolerance = 0.05,
#                          preserveTopology = TRUE)
# writeVector(ipbesred, "../Data_geo/ipbes_regions_subregions/ipbes_reduced.shp")
ipbesred <- vect(ipbes_shp[1])
plot(ipbesred)
```

# 2. Cube/entity metadata

For `ebvcube` models, PLUM (id102) cubes carry three metrics per scenario
(protected / unprotected / total); other EBVcube models may carry only
one. The per-model `metric_filter` block in the YAML declares which metric
path(s) to consume — everything downstream sees only the filtered set.
Setting `metric_filter: ~` in the YAML disables filtering for that model.

For `geotiff_dir` models there is no EBVcube metadata to read — entity
names, scenarios and years come directly from the YAML, and the spatial
template is read from the first available GeoTIFF.

```{r metadata}
if (input_type == "ebvcube") {

  cube_props     <- ebv_properties(cube_path,    verbose = FALSE)
  cube_paths_all <- ebv_datacubepaths(cube_path, verbose = FALSE)

  # Apply per-model metric filter
  cube_paths <- if (!is.null(mcfg$metric_filter$path_pattern)) {
    cube_paths_all[grepl(mcfg$metric_filter$path_pattern,
                         cube_paths_all$datacubepaths), ]
  } else if (!is.null(mcfg$metric_filter$keep)) {
    cube_paths_all[cube_paths_all$metric_names %in%
                   mcfg$metric_filter$keep, ]
  } else {
    cube_paths_all
  }

  stopifnot(nrow(cube_paths) >= 1)

  cat("Model:     ", model_label, "\n")
  cat("Metrics retained after filter:\n")
  print(unique(cube_paths$metric_names))
  cat("Scenarios retained after filter:\n")
  print(unique(cube_paths$scenario_names))

  cat("Entities:\n");   print(cube_props@general$entity_names)
  cat("Timesteps:\n");  print(cube_props@temporal$dates)
  cat("EPSG:      ",    cube_props@spatial$epsg,       "\n")
  cat("Resolution:",    cube_props@spatial$resolution, "\n")
  cat("Extent:    ",    cube_props@spatial$extent,     "\n")

  print(cube_paths)

} else if (input_type == "geotiff_dir") {

  # TEMPORARY (see TODO in section 0): metadata assembled from YAML +
  # a sample GeoTIFF instead of ebv_properties()/ebv_datacubepaths().
  # Entity order here MUST match the numbering of the entity_NN files on
  # disk (entity_01, entity_02, ...) — i.e. entry order in the YAML's
  # `entities:` map for this model is the contract, same as the original
  # PLUM_Alexander notebook's hardcoded entity_names vector.
  entity_names_geo <- names(mcfg$entities)

  template_tif <- file.path(
    tif_root, mcfg$scenarios[1], as.character(mcfg$years[1]),
    sprintf(mcfg$entity_file_pattern, 1)
  )
  stopifnot(file.exists(template_tif))
  geotiff_template <- rast(template_tif)

  cat("Model:     ", model_label, "\n")
  cat("Scenarios available:\n"); print(mcfg$scenarios)
  cat("Years available:\n");     print(mcfg$years)
  cat("Entities (file order):\n"); print(entity_names_geo)
  cat("Spatial template:\n");    print(geotiff_template)
}
```

# 3. Category map for the active model

Original entity names differ between models, so each gets its own
mapping from native classes to the six target classes (`Urban`,
`Cropland`, `Secondary_vegetation`, `Permanent_crops`, `Pasture`,
`Primary_vegetation`). Entities mapped to `~` (YAML null → R `NA`) are
excluded from the habitat-area denominator before normalization
(e.g. `Barren` in PLUM, since polar/desert/rock pixels would otherwise
distort the regional richness estimates). This section is identical for
both input backends — it only depends on `mcfg$entities`.

```{r category-map}
category_map <- unlist(mcfg$entities)

# Normalize potential string-form nulls to true NA
category_map[category_map %in% c("", "NA")] <- NA_character_

if (length(category_map) == 0) {
  warning("category_map for '", model_label,
          "' is empty — fill it in the YAML before running sections 4-5.")
}

# Entities excluded from the recategorization sum
excluded_entities <- names(category_map)[is.na(category_map)]
if (length(excluded_entities) > 0) {
  cat("Excluded from habitat denominator (NA-mapped):\n")
  print(excluded_entities)
}

# Active map: only entries with a defined target class.
# This is the ONLY map used downstream for recategorization —
# NA-mapped entities are never summed into any target class.
category_map_active <- category_map[!is.na(category_map)]
new_categories       <- unique(category_map_active)

cat("Target classes produced for", model_label, ":\n")
print(new_categories)

category_map
```

# 4. Changes in area

Both the percentage-point and km² outputs below come from a single
fraction-read pass per entity — fractions are read once (section 4.1),
then reused to derive km² (section 4.2) by multiplying by a
latitude-varying pixel-area raster. No entity is read from disk/cube more
than once, regardless of backend.

## 4.1 Reading fractions and percentage-point change per decade

Per-decade change is computed directly from the fraction layers
(`(y_t1 - y_t0) / n_decades`), with `y` being the proportion (%) of a
pixel covered by a target class.

```{r}
# Select scenario by NAME, not row index — robust across models with
# different scenario counts/labels. Change scenario_choice to switch.
# For ebvcube models this must match a value in cube_paths$scenario_names;
# for geotiff_dir models it must match one of mcfg$scenarios.
scenario_choice <- if (input_type == "ebvcube") "SSP2 RCP 6.0" else mcfg$scenarios[1]

t0_label <- "2020"
t1_label <- "2050"

if (input_type == "ebvcube") {

  scenario_row <- cube_paths[cube_paths$scenario_names == scenario_choice, ]
  stopifnot(nrow(scenario_row) == 1)

  dc_path <- scenario_row$datacubepaths

  cat("Processing scenario:", scenario_row$scenario_names,
      "| metric:", scenario_row$metric_names, "\n")

  # Metadata (shared by both 4.1 and 4.2)
  entity_names        <- cube_props@general$entity_names
  safe_entity_names   <- gsub("[[:space:]/]+", "_", entity_names)
  timestep_labels_all <- format(as.Date(cube_props@temporal$dates), "%Y")
  n_entities          <- length(entity_names)

  t0_idx   <- which(timestep_labels_all == t0_label)
  t1_idx   <- which(timestep_labels_all == t1_label)
  keep_idx <- c(t0_idx, t1_idx)
  timestep_labels <- timestep_labels_all[keep_idx]

  # Read all original entities for t0 and t1 — single read, reused below
  lu_frac <- vector("list", n_entities)
  names(lu_frac) <- safe_entity_names

  for (e in seq_len(n_entities)) {
    cat(sprintf("[%d/%d] %s\n", e, n_entities, entity_names[e]))
    frac <- ebv_read(
      filepath     = cube_path,
      datacubepath = dc_path,
      entity       = e,
      timestep     = keep_idx,
      ignore_RAM   = TRUE,
      verbose      = FALSE
    )
    names(frac) <- paste0(safe_entity_names[e], "_", timestep_labels)
    lu_frac[[safe_entity_names[e]]] <- frac
  }

} else if (input_type == "geotiff_dir") {

  # TEMPORARY (see TODO in section 0): reads per-entity GeoTIFFs directly
  # instead of ebv_read(). Replace with an EBVcube read once available.
  stopifnot(scenario_choice %in% mcfg$scenarios)
  stopifnot(t0_label %in% as.character(mcfg$years))
  stopifnot(t1_label %in% as.character(mcfg$years))

  cat("Processing scenario:", scenario_choice, "\n")

  entity_names      <- entity_names_geo
  safe_entity_names <- gsub("[[:space:]/]+", "_", entity_names)
  n_entities        <- length(entity_names)
  timestep_labels   <- c(t0_label, t1_label)

  lu_frac <- vector("list", n_entities)
  names(lu_frac) <- safe_entity_names

  for (e in seq_len(n_entities)) {
    cat(sprintf("[%d/%d] %s\n", e, n_entities, entity_names[e]))

    tif_t0 <- file.path(tif_root, scenario_choice, t0_label,
                         sprintf(mcfg$entity_file_pattern, e))
    tif_t1 <- file.path(tif_root, scenario_choice, t1_label,
                         sprintf(mcfg$entity_file_pattern, e))
    stopifnot(file.exists(tif_t0), file.exists(tif_t1))

    frac <- c(rast(tif_t0), rast(tif_t1))
    names(frac) <- paste0(safe_entity_names[e], "_", timestep_labels)
    lu_frac[[safe_entity_names[e]]] <- frac
  }
}

cat("Done.", model_label, "-", n_entities, "entities x 2 timesteps.\n")

# Verify entity coverage of the category map
unmapped <- setdiff(names(lu_frac), names(category_map))
if (length(unmapped) > 0) {
  stop("Entities present in cube but missing from category_map: ",
       paste(unmapped, collapse = ", "),
       "\n -> add them to config/category_map.yaml under models$",
       model_key, "$entities")
}

# Recategorise: sum fractions across entities sharing a new category.
# NA-mapped entities are already excluded — category_map_active and
# new_categories come from section 3.
lu_frac_recat <- lapply(new_categories, function(new_cat) {
  old_entities  <- names(category_map_active)[category_map_active == new_cat]
  summed        <- Reduce(`+`, lu_frac[old_entities])
  names(summed) <- paste0(new_cat, "_", timestep_labels)
  summed
}) |> setNames(new_categories)

cat("Recategorised into", length(new_categories), "classes:\n")
print(new_categories)

# Percentage-point change per decade
n_decades <- (as.numeric(t1_label) - as.numeric(t0_label)) / 10

lu_pct_change_per_decade <- lapply(names(lu_frac_recat), function(lu) {
  r  <- lu_frac_recat[[lu]]
  y0 <- r[[paste0(lu, "_", t0_label)]]
  y1 <- r[[paste0(lu, "_", t1_label)]]
  change_per_decade <- (y1 - y0) / n_decades
  names(change_per_decade) <-
    paste0(lu, "_pctchange_per_decade_", t0_label, "_", t1_label)
  change_per_decade
}) |> setNames(names(lu_frac_recat))

cat("Percentage-point change per decade computed for", model_label, ":\n")
print(names(lu_pct_change_per_decade))

plot(lu_pct_change_per_decade$Primary_vegetation,
     main = paste(model_label, "- Primary_vegetation pct change/decade"))
```

```{r}
pct_change_stack <- rast(lu_pct_change_per_decade)
plot(pct_change_stack, main = paste(model_label, names(pct_change_stack)))
```

## 4.2 Changes in km²

Fractions are converted to km² using a latitude-varying pixel-area
raster (`terra::cellSize`), reusing the `lu_frac` already read in 4.1 —
no second read pass, regardless of backend.

```{r}
# Pixel-area raster (latitude-varying), built once
if (input_type == "ebvcube") {
  pixel_area_km2 <- cellSize(
    rast(
      extent     = ext(cube_props@spatial$extent),
      resolution = cube_props@spatial$resolution,
      crs        = paste0("EPSG:", cube_props@spatial$epsg)
    ),
    unit = "km"
  )
} else if (input_type == "geotiff_dir") {
  # TEMPORARY (see TODO in section 0)
  pixel_area_km2 <- cellSize(geotiff_template, unit = "km")
}

# Convert each already-read entity's fractions to km²
lu_area_km2 <- lapply(lu_frac, function(r) r * pixel_area_km2)
names(lu_area_km2) <- names(lu_frac)

cat("Converted to km² for", model_label, "-", length(lu_area_km2), "entities.\n")

# Recategorise: sum km² across entities sharing a new category
lu_area_km2_recat <- lapply(new_categories, function(new_cat) {
  old_entities  <- names(category_map_active)[category_map_active == new_cat]
  summed        <- Reduce(`+`, lu_area_km2[old_entities])
  names(summed) <- paste0(new_cat, "_", timestep_labels)
  summed
}) |> setNames(new_categories)

cat("Reclassified into", length(new_categories), "categories for",
    model_label, ":\n")
print(new_categories)
```

```{r}
lu_area_stack <- rast(lu_area_km2_recat)
plot(lu_area_stack, main = paste(model_label, names(lu_area_stack)))

# Primary_vegetation: t0, t1, and the difference (t1 - t0)
pv_t0 <- lu_area_km2_recat$Primary_vegetation[[paste0("Primary_vegetation_", t0_label)]]
pv_t1 <- lu_area_km2_recat$Primary_vegetation[[paste0("Primary_vegetation_", t1_label)]]
pv_diff <- pv_t1 - pv_t0
names(pv_diff) <- paste0("Primary_vegetation_diff_", t0_label, "_", t1_label)

pv_compare <- c(pv_t0, pv_t1, pv_diff)

plot(pv_compare,
     main = paste(model_label, "-", names(pv_compare)))
```
# Plots

Two kind o plots are developed, one with continuous data and the other with binned data.
```{r}
# ── 4.3 Map: percentage-point change per decade, choropleth style ───────────
# Two alternatives — pick whichever fits the deliverable:
#   A) Continuous scale, fixed limits (consistent with earlier decision)
#   B) Binned/quantile scale styled after the reference figure (asymmetric breaks)

#loading coastlines:

library(rnaturalearth)

# Get country/coastline boundaries
coastline <- ne_coastline(scale = "medium", returnclass = "sf")
coastline <- st_make_valid(coastline)



# Use any single recategorised entity here, e.g. Primary_vegetation
pv_pct_change <- lu_pct_change_per_decade$Primary_vegetation

# Convert to a long data frame for ggplot
pv_df <- as.data.frame(pv_pct_change, xy = TRUE, na.rm = TRUE)
names(pv_df)[3] <- "pct_change"

# ── Alternative A: continuous scale (consistent with our earlier decision) ──

ggplot(pv_df) +
  geom_tile(aes(x = x, y = y, fill = pct_change)) +
  geom_sf(data = coastline, color = "grey30", linewidth = 0.1, inherit.aes = FALSE) +
  scale_fill_distiller(
    palette  = "Reds",
    direction = -1,
    limits   = c(min(pv_df$pct_change, na.rm = TRUE), 0),
    oob      = scales::squish,
    na.value = NA,
    name     = "pp / decade"
  ) +
  coord_sf(expand = FALSE, xlim = c(-180, 180), ylim = c(-90, 90)) +
  labs(title = paste0("Δ ", t0_label, "–", t1_label)) +
  theme_void() +
  theme(
    legend.position   = "bottom",
    legend.key.width  = unit(2, "cm"),
    legend.key.height = unit(0.3, "cm")
  )

# ── Alternative B: binned/quantile scale, styled after the reference figure ─

breaks_b <- c(min(pv_df$pct_change, na.rm = TRUE), -0.2, -0.1, -0.05, -0.02, -0.01, 0)

reds_colors <- RColorBrewer::brewer.pal(9, "Reds")  # adjust n to match your number of bins
reds_colors <- rev(reds_colors)  
reds_colors[length(reds_colors)] <- "#FFF7BC" 

b <- ggplot(pv_df) +
  geom_tile(aes(x = x, y = y, fill = pct_change)) +
  geom_sf(data = coastline, color = "grey30", linewidth = 0.1, inherit.aes = FALSE) +
  
  scale_fill_stepsn(
  colors   = reds_colors,
  breaks   = breaks_b,
  labels   = round(sort(breaks_b), 2),
  na.value = NA,
  name     = "percentage points per decade"
  ) +
  
#  scale_fill_fermenter(
#    palette  = "Reds",
#    direction = -1,
#    breaks   = breaks_b,
#    labels = scales::label_number(accuracy = 0.01)(sort(breaks_b)),   # round to 2 decimal places
#    na.value = NA,
#    name     = "percentage points per decade"
#  ) +
  coord_sf(expand = FALSE, xlim = c(-180, 180), ylim = c(-90, 90)) +
  labs(title = paste0("Δ ", t0_label, "–", t1_label, " SSP2 RCP 6.0")) +
  theme_void() +
  theme(
    legend.position   = "bottom",
    legend.key.width  = unit(1.4, "cm"),
    legend.key.height = unit(0.3, "cm")
  )

b
```
diagnose for the distribution of points in the plot?

```{r}
#breaks_b <- c(min(pv_df$pct_change, na.rm = TRUE), -3.08, -1.16, -0.32, -0.30, -0.2, -0.1, -0.05, -0.01, 0)

#table(cut(pv_df$pct_change, breaks = sort(unique(breaks_b))))
```
NOTE: to apply to SSP2 RCP6.

The problem is a mismatch between the number of colors and the number of bins. You have 6 breaks, which creates 5 bins (plus the out-of-range bin below the minimum), but brewer.pal(9, "Reds") gives you 9 colors. scale_fill_stepsn() interpolates across all 9 colors to fill however many bins exist — so the yellow (#FFF7BC) at position 9 gets compressed into a barely-visible sliver or disappears entirely because the interpolation never reaches the last stop.
Fix — match the number of colors to the number of bins:

```{r}
breaks_b <- c(min(pv_df$pct_change, na.rm = TRUE), -0.2, -0.1, -0.05, -0.02, -0.01, 0)
table(cut(pv_df$pct_change, breaks = sort(unique(breaks_b))))
```

```{r}
breaks_b <- c(-3.08, -1.16, -0.32, -0.30, -0.2, -0.1, -0.05, -0.01, 0)

n_colors <- length(breaks_b) + 1   # = 10

reds_colors <- RColorBrewer::brewer.pal(n_colors, "Reds")
reds_colors <- rev(reds_colors)
reds_colors[length(reds_colors)] <- "#FFF7BC"

ggplot(pv_df) +
  geom_tile(aes(x = x, y = y, fill = pct_change)) +
  geom_sf(data = coastline, color = "grey30", linewidth = 0.1, inherit.aes = FALSE) +
  scale_fill_stepsn(
    colors   = reds_colors,
    breaks   = breaks_b,
    limits   = c(min(pv_df$pct_change, na.rm = TRUE), 0),
    labels   = round(breaks_b, 2),
    na.value = NA,
    name     = "percentage points per decade"
  ) +
  coord_sf(expand = FALSE, xlim = c(-180, 180), ylim = c(-90, 90)) +
  labs(title = paste0("Δ ", t0_label, "–", t1_label, " SSP2 RCP 6.0")) +
  theme_void() +
  theme(
    legend.position   = "bottom",
    legend.key.width  = unit(1.4, "cm"),
    legend.key.height = unit(0.3, "cm")
  )
```


## Fraction of natural land in 2020

```{r}
# ── Distribution of Primary vegetation in 2020 (%) ───────────────────────────
pv_pct_2020 <- lu_frac_recat$Primary_vegetation[[paste0("Primary_vegetation_", t0_label)]] * 100
names(pv_pct_2020) <- "Primary_vegetation_pct_2020"

pv_pct_df <- as.data.frame(pv_pct_2020, xy = TRUE, na.rm = TRUE)
names(pv_pct_df)[3] <- "pct"

# Quantile-based breaks — replaces the hardcoded reference-figure values
probs_dist <- c(0, 0.05, 0.15, 0.3, 0.45, 0.55, 0.7, 0.85, 0.95, 1)
breaks_dist <- quantile(pv_pct_df$pct, probs = probs_dist, na.rm = TRUE)
breaks_dist <- unname(breaks_dist)

# Diagnostic: check bin counts before plotting
print(table(cut(pv_pct_df$pct, breaks = sort(unique(breaks_dist)))))

# Diverging brown -> blue palette (BrBG)
brbg_colors <- RColorBrewer::brewer.pal(length(unique(breaks_dist)) - 1, "BrBG")

a <- ggplot(pv_pct_df) +
  geom_tile(aes(x = x, y = y, fill = pct)) +
  geom_sf(data = coastline, color = "grey30", linewidth = 0.1, inherit.aes = FALSE) +
  scale_fill_stepsn(
    colors   = brbg_colors,
    breaks   = breaks_dist,
    labels   = round(breaks_dist, 1),
    limits   = c(0, 100),
    na.value = NA,
    name     = paste0("distribution of land in ", t0_label, " (%)")
  ) +
  coord_sf(expand = FALSE, xlim = c(-180, 180), ylim = c(-90, 90)) +
  labs(title = t0_label) +
  theme_void() +
  theme(
    legend.position    = "bottom",
    legend.key.width   = unit(1.4, "cm"),
    legend.key.height  = unit(0.3, "cm"),
    plot.title         = element_text(size = 14, face = "bold", hjust = 0)
  )

a
```
```{r}
# Quantile-based breaks — each bin gets roughly equal pixel count
probs_dist <- c(0, 0.05, 0.15, 0.3, 0.45, 0.55, 0.7, 0.85, 0.95, 1)
breaks_dist <- quantile(pv_pct_df$pct, probs = probs_dist, na.rm = TRUE)
breaks_dist <- unname(breaks_dist)

breaks_dist

# Re-check the distribution across these new breaks
table(cut(pv_pct_df$pct, breaks = sort(unique(breaks_dist))))
```
