---
title: "4_land_use_changes"
output: html_notebook
---

```{r setup, message=FALSE}
library(ebvcube)
library(tidyterra)
library(terra)
library(tidyverse)
library(graticule)
library(sf)
library(rworldmap)
library(ggplot2)
library(stringr)

#setting the working directory
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
```

# 1. Paths

Here we call the EBVcube file with GLOBIO. For the purpose of this exploratory analysis I need to simplify the ipbes vector file, so that it runs faster. To provide a fast analysis, I have set a tolerance parameter of `0.05` meaning that within a 0.05 degree grid, every vertice would be flattened into a single straight line. Given that the rasters we are working with are 0.5, this does not create any practical problem.

```{r}
root_LU_cubes <- "../Data_geo/"

# EBVcube NetCDFs
dc <- list(
  globio_lu_05 = "../Data_geo/land_use/GLOBIO/cubes/alkemade_ecostr_id103_20260216.nc",
  plum_lu_05   = "../Data_geo/land_use/PLUM/cubes/alexander_ecostr_id102_20260320.nc",
  scenmip_0.25 = "../Data_geo/land_use/ScenMIP7-LUH3/cubes/chini_ecostr_id105_20260325.nc"
)

# IPBES regions shapefile — used in regional aggregation (next step)
ipbes_shp <- list.files(root_LU_cubes, pattern = "\\.shp$",
                        full.names = TRUE, ignore.case = TRUE, recursive = TRUE)
ipbes_shp

#Downscaling the IPBES shape file directly from the original file
#ipbes <- vect(ipbes_shp)
#ipbesred <- simplifyGeom(ipbes, tolerance = 0.05, preserveTopology = TRUE)
#writeVector(ipbesred, "../Data_geo/ipbes_regions_subregions/ipbes_reduced.shp")

# calling the reduced shape (time saver)
ipbesred <- vect(ipbes_shp[1])
plot(ipbesred)
```

# 2. GLOBIO cube metadata

```{r globio-metadata}
globio05       <- ebv_properties(dc$plum_lu_05, verbose = FALSE)
globio05_paths <- ebv_datacubepaths(dc$plum_lu_05, verbose = FALSE)

# Available scenarios: NfN [1,1] | NaC [2,1] | NfS [3,1] | BAU [4,1]
print(globio05_paths)

cat("Entities:\n");   print(globio05@general$entity_names)
cat("Timesteps:\n");  print(globio05@temporal$dates)
cat("EPSG:      ",    globio05@spatial$epsg,       "\n")
cat("Resolution:",    globio05@spatial$resolution, "\n")
cat("Extent:    ",    globio05@spatial$extent,     "\n")
```

# 3. Double checking changes in Area

## 3.1 Changes in absolute relative percentages

```{r}
# ── Percentage-point change per decade, 2020 → 2070, directly from fractions ─
# Formula: ((y_t1 - y_t0) / (t1 - t0)) * 10
# y = % of pixel covered by a land-use type, read directly from the EBVcube
# (no conversion to km² — these are fraction/percentage values throughout)

# ── 1. Cube path for the scenario ────────────────────────────────────────────

dc_path <- globio05_paths[1, 1]   # NfN — change row index to switch scenario

# ── 2. Read all 12 original entities, both timesteps (2020 and 2070) ────────

entity_names      <- globio05@general$entity_names
safe_entity_names <- gsub("[[:space:]/]+", "_", entity_names)
timestep_labels   <- format(as.Date(globio05@temporal$dates), "%Y")
n_entities        <- length(entity_names)

t0_label <- "2020"
t1_label <- "2070"

t0_idx <- which(timestep_labels == t0_label)
t1_idx <- which(timestep_labels == t1_label)

globio_frac <- vector("list", n_entities)
names(globio_frac) <- safe_entity_names

for (e in seq_len(n_entities)) {

  cat(sprintf("[%d/%d] %s\n", e, n_entities, entity_names[e]))

  frac <- ebv_read(
    filepath     = dc$globio_lu_05,
    datacubepath = dc_path,
    entity       = e,
    timestep     = c(t0_idx, t1_idx),
    ignore_RAM   = TRUE,
    verbose      = FALSE
  )

  names(frac) <- paste0(safe_entity_names[e], "_", c(t0_label, t1_label))
  globio_frac[[safe_entity_names[e]]] <- frac
}

cat("Done.", n_entities, "entities ×", length(c(t0_idx, t1_idx)), "timesteps.\n")

# ── 3. Recategorise: sum fractions across entities sharing a new category ──

category_map <- c(
  Urban                           = "Urban",
  Cropland                        = "Cropland",
  Potential_secondary_vegetation  = "Secondary_vegetation",
  Agroforestry                    = "Permanent_crops",
  Pasture                         = "Pasture",
  Rangeland                       = "Pasture",
  Forestry                        = "Secondary_vegetation",
  Secondary_vegetation            = "Secondary_vegetation",
  Unmanaged_forest                = "Primary_vegetation",
  Shrubland_Grassland             = "Primary_vegetation",
  Other                           = "Primary_vegetation",
  Bare_Areas                      = "Primary_vegetation"
)

stopifnot(all(names(globio_frac) %in% names(category_map)))

new_categories <- unique(category_map)

globio_frac_recat <- lapply(new_categories, function(new_cat) {
  old_entities  <- names(category_map)[category_map == new_cat]
  summed        <- Reduce(`+`, globio_frac[old_entities])
  names(summed) <- paste0(new_cat, "_", c(t0_label, t1_label))
  summed
}) |> setNames(new_categories)

cat("Recategorised into", length(new_categories), "classes:\n")
print(new_categories)

# ── 4. Percentage-point change per decade, per recategorised entity ─────────

n_decades <- (as.numeric(t1_label) - as.numeric(t0_label)) / 10

globio_pct_change_per_decade <- lapply(names(globio_frac_recat), function(lu) {

  r  <- globio_frac_recat[[lu]]
  y0 <- r[[paste0(lu, "_", t0_label)]]
  y1 <- r[[paste0(lu, "_", t1_label)]]

  change_per_decade <- ((y1 - y0) / n_decades)
  names(change_per_decade) <- paste0(lu, "_pctchange_per_decade_", t0_label, "_", t1_label)
  change_per_decade

}) |> setNames(names(globio_frac_recat))

cat("Percentage-point change per decade computed for:\n")
print(names(globio_pct_change_per_decade))

# Quick check
plot(globio_pct_change_per_decade$Secondary_vegetation)
```

```{r}
pct_change_stack <- rast(globio_pct_change_per_decade)
plot(pct_change_stack)
```

## 3.2 Changes in km²

```{r}
dc_path <- globio05_paths[1, 1]   

# Pixel-area raster — latitude-varying, built once
pixel_area_km2 <- cellSize(
  rast(extent = ext(-180, 180, -90, 90), resolution = 0.5, crs = "EPSG:4326"),
  unit = "km"
)

# Metadata for the loop
entity_names      <- globio05@general$entity_names
safe_entity_names <- gsub("[[:space:]/]+", "_", entity_names)
timestep_labels_all <- format(as.Date(globio05@temporal$dates), "%Y")
n_entities        <- length(entity_names)

# Restrict to 2020 and 2070 only
t0_label <- "2020"
t1_label <- "2070"
keep_idx <- which(timestep_labels_all %in% c(t0_label, t1_label))
timestep_labels <- timestep_labels_all[keep_idx]
n_timesteps     <- length(timestep_labels)

# Loop: read only 2020 & 2070 per entity, convert fraction → km², name layers
globio_area_km2 <- vector("list", n_entities)
names(globio_area_km2) <- safe_entity_names

for (e in seq_len(n_entities)) {

  cat(sprintf("[%d/%d] %s\n", e, n_entities, entity_names[e]))

  frac <- ebv_read(
    filepath     = dc$globio_lu_05,
    datacubepath = dc_path,
    entity       = e,
    timestep     = keep_idx,
    ignore_RAM   = TRUE,
    verbose      = FALSE
  )

  area        <- frac * pixel_area_km2
  names(area) <- paste0(safe_entity_names[e], "_", timestep_labels)

  globio_area_km2[[safe_entity_names[e]]] <- area
}

cat("Done.", n_entities, "entities ×", n_timesteps, "timesteps.\n")
```

```{r}
# mapping old categories to the LUH3 reclass

category_map <- c(
  Urban                           = "Urban",
  Cropland                        = "Cropland",
  Potential_secondary_vegetation  = "Secondary_vegetation",
  Agroforestry                    = "Permanent_crops",
  Pasture                         = "Pasture",
  Rangeland                       = "Pasture",
  Forestry                        = "Secondary_vegetation",
  Secondary_vegetation            = "Secondary_vegetation",
  Unmanaged_forest                = "Primary_vegetation",
  Shrubland_Grassland             = "Primary_vegetation",
  Other                           = "Primary_vegetation",
  Bare_Areas                      = "Primary_vegetation"
)

stopifnot(all(names(globio_area_km2) %in% names(category_map)))
```

Applying these changes to all other rasters

```{r}
new_categories <- unique(category_map)

globio_area_km2_recat <- lapply(new_categories, function(new_cat) {
  old_entities <- names(category_map)[category_map == new_cat]
  # Sum the corresponding rasters (same layer structure: 2 timesteps each — 2020 & 2070)
  rasters_to_sum <- globio_area_km2[old_entities]
  summed <- Reduce(`+`, rasters_to_sum)
  names(summed) <- paste0(new_cat, "_", timestep_labels)
  summed
})

names(globio_area_km2_recat) <- new_categories

cat("Reclassified into", length(new_categories), "categories:\n")
print(new_categories)
```

```{r}
globio_area_stack <- rast(globio_area_km2_recat)
plot(globio_area_stack)

plot(globio_area_km2_recat$Primary_vegetation)
```
