---
title: "RDelta Comparison Across BES-SIM 2 Scenarios — GLOBIO"
author: "Andrés Camilo Mármol Guijarro"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 12,
  fig.height = 7
)
```

```{r libraries}
library(tidyverse)
library(here)
library(scales)
library(patchwork)
```

---

# 1 · Load and prepare data

```{r load-data}
rdelta_dir <- "/Volumes/AM_SSD/BES-SIM2_cSAR/cSAR_BES-SIM_2/BES_results/results_biodiversity/cSAR_iDiv/"

rdelta_paths <- list.files(
  rdelta_dir,
  pattern   = "^rdelta_.*\\.csv$",
  full.names = TRUE
)

# Keep only the most recent run per scenario (2020-2050 prefix in name, dated 23062026)
# If you want all files, remove this filter
rdelta_paths_2050 <- rdelta_paths[grepl("23062026", rdelta_paths)]

# Parse scenario label from filename
parse_scenario <- function(path) {
  fname <- basename(path)
  # extract token between first underscore and _GLOBIO
  sub("^rdelta_(.+)_GLOBIO.*$", "\\1", fname)
}

# Read and bind, tagging each row with its scenario
rdelta_raw <- map_dfr(rdelta_paths_2050, function(p) {
  read_csv(p, show_col_types = FALSE) |>
    mutate(scenario = parse_scenario(p))
})

# Clean up: drop row-index column, parse year_comparison to numeric endpoint
rdelta <- rdelta_raw |>
  select(-any_of("...1")) |>
  mutate(
    # ── normalise taxon label across files ──────────────────────────────────
    taxon = case_when(
      str_detect(taxon, regex("non.?forest", ignore_case = TRUE)) ~ "Non-forest",
      str_detect(taxon, regex("^forest$",    ignore_case = TRUE)) ~ "Forest",
      .default = taxon
    ),
    # ── factor levels ───────────────────────────────────────────────────────
    year_end = as.integer(sub("2020_", "", year_comparison)),
    taxon    = factor(taxon, levels = c("Forest", "Non-forest")),
    scenario = factor(scenario, levels = c("NfN", "NfS", "NaC", "SSP2RCP6"))
  )

glimpse(rdelta)
```

```{r helpers}
# Shared colour palette — one colour per scenario
scenario_cols <- c(
  NfN      = "#1b7837",   # deep green  (Nature for Nature)
  NfS      = "#762a83",   # purple      (Nature for Society)
  NaC      = "#e08214",   # amber       (Nature as Culture)
  SSP2RCP6 = "#d73027"    # red         (BAU / SSP2)
)

# Horizontal reference line + theme used across all plots
base_theme <- function() {
  theme_bw(base_size = 12) +
  theme(
    strip.background = element_rect(fill = "grey92"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )
}
```

---

# 2 · IPBES Regions

```{r region-forest}
regions <- rdelta |> filter(zone_type == "Region")

# ── 2a  Forest ──────────────────────────────────────────────────────────────
p_reg_forest <- regions |>
  filter(taxon == "Forest") |>
  ggplot(aes(x = year_end, y = RDelta, colour = scenario, group = scenario)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.4) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  facet_wrap(~zone, ncol = 3) +
  scale_colour_manual(values = scenario_cols, name = "Scenario") +
  scale_x_continuous(breaks = c(2030, 2050, 2070)) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title    = "Forest species: relative richness change (RΔ) by IPBES region",
    subtitle = "Relative to 2020 baseline · GLOBIO",
    x        = "Year",
    y        = "RΔ"
  ) +
  base_theme()

p_reg_forest
```

```{r region-nonforest}
# ── 2b  Non-forest ──────────────────────────────────────────────────────────
p_reg_nonforest <- regions |>
  filter(taxon == "Non-forest") |>
  ggplot(aes(x = year_end, y = RDelta, colour = scenario, group = scenario)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.4) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  facet_wrap(~zone, ncol = 3) +
  scale_colour_manual(values = scenario_cols, name = "Scenario") +
  scale_x_continuous(breaks = c(2030, 2050, 2070)) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title    = "Non-forest species: relative richness change (RΔ) by IPBES region",
    subtitle = "Relative to 2020 baseline · GLOBIO",
    x        = "Year",
    y        = "RΔ"
  ) +
  base_theme()

p_reg_nonforest
```

```{r region-combined, fig.height=5}
# ── 2c  Both taxa side-by-side per region (2050 snapshot) ───────────────────
p_reg_bar <- regions |>
  filter(year_end == 2050) |>
  ggplot(aes(x = zone, y = RDelta, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40", linewidth = 0.4) +
  facet_wrap(~taxon, ncol = 2) +
  scale_fill_manual(values = scenario_cols, name = "Scenario") +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title    = "RΔ at 2050 by IPBES region and taxon group",
    subtitle = "GLOBIO · relative to 2020",
    x        = NULL,
    y        = "RΔ"
  ) +
  base_theme() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

p_reg_bar
```

```{r}
p_reg_bar <- regions |>
  filter(year_end == 2050) |>
  ggplot(aes(x = reorder(zone, RDelta), y = RDelta, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40", linewidth = 0.4) +
  facet_wrap(~taxon, ncol = 2) +
  scale_fill_manual(values = scenario_cols, name = "Scenario") +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  coord_flip() +
  labs(
    title    = "RΔ at 2050 by IPBES region and taxon group",
    subtitle = "GLOBIO · relative to 2020 · ordered by NfN RΔ",
    x        = NULL,
    y        = "RΔ"
  ) +
  base_theme()
p_reg_bar
```

---

# 3 · IPBES Subregions

```{r subregion-forest, fig.height=10}
subregions <- rdelta |> filter(zone_type == "Sub_Region")

# ── 3a  Forest ──────────────────────────────────────────────────────────────
p_sub_forest <- subregions |>
  filter(taxon == "Forest") |>
  ggplot(aes(x = year_end, y = RDelta, colour = scenario, group = scenario)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.4) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~zone, ncol = 4) +
  scale_colour_manual(values = scenario_cols, name = "Scenario") +
  scale_x_continuous(breaks = c(2030, 2050, 2070)) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title    = "Forest species: RΔ by IPBES subregion",
    subtitle = "Relative to 2020 baseline · GLOBIO",
    x        = "Year",
    y        = "RΔ"
  ) +
  base_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))

p_sub_forest
```

```{r subregion-nonforest, fig.height=10}
# ── 3b  Non-forest ──────────────────────────────────────────────────────────
p_sub_nonforest <- subregions |>
  filter(taxon == "Non-forest") |>
  ggplot(aes(x = year_end, y = RDelta, colour = scenario, group = scenario)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.4) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~zone, ncol = 4) +
  scale_colour_manual(values = scenario_cols, name = "Scenario") +
  scale_x_continuous(breaks = c(2030, 2050, 2070)) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title    = "Non-forest species: RΔ by IPBES subregion",
    subtitle = "Relative to 2020 baseline · GLOBIO",
    x        = "Year",
    y        = "RΔ"
  ) +
  base_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))

p_sub_nonforest
```

```{r subregion-bar, fig.height=9}
# ── 3c  Both taxa · 2050 snapshot bar chart ─────────────────────────────────
p_sub_bar <- subregions |>
  filter(year_end == 2050) |>
  ggplot(aes(x = reorder(zone, RDelta), y = RDelta, fill = scenario)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40", linewidth = 0.4) +
  facet_wrap(~taxon, ncol = 2) +
  scale_fill_manual(values = scenario_cols, name = "Scenario") +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  coord_flip() +
  labs(
    title    = "RΔ at 2050 by IPBES subregion and taxon group",
    subtitle = "GLOBIO · relative to 2020 · ordered by NfN RΔ",
    x        = NULL,
    y        = "RΔ"
  ) +
  base_theme()

p_sub_bar
```

---

# 4 · Export plots


out_dir <- here("BES_results", "results_biodiversity", "cSAR_iDiv", "figures")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

ggsave(file.path(out_dir, "rdelta_region_forest_timeseries.png"),    p_reg_forest,    width = 14, height = 8,  dpi = 300)
ggsave(file.path(out_dir, "rdelta_region_nonforest_timeseries.png"), p_reg_nonforest, width = 14, height = 8,  dpi = 300)
ggsave(file.path(out_dir, "rdelta_region_bar_2050.png"),             p_reg_bar,       width = 14, height = 6,  dpi = 300)
ggsave(file.path(out_dir, "rdelta_subregion_forest_timeseries.png"), p_sub_forest,    width = 16, height = 10, dpi = 300)
ggsave(file.path(out_dir, "rdelta_subregion_nonforest_timeseries.png"), p_sub_nonforest, width = 16, height = 10, dpi = 300)
ggsave(file.path(out_dir, "rdelta_subregion_bar_2050.png"),          p_sub_bar,       width = 14, height = 10, dpi = 300)

message("Figures saved to: ", out_dir)
```
